\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

This thesis presents a Visual Studio Code extension for visual analysis of programs written in Stella, an educational statically typed programming language. The problem addressed in this work is that standard editor diagnostics expose the final result of language analysis but do not show the intermediate structures used during parsing and type checking. The developed extension makes five kinds of information available inside the editor: an abstract syntax tree, a scope and typing-context view, a type derivation tree, a reverse type-error trace, and a type dependency graph. The implementation uses TypeScript, the Visual Studio Code Extension API, Langium, and custom Language Server Protocol requests. The client and server are separated so that semantic analysis remains on the language-server side, while editor-specific rendering is implemented on the client side. The prototype was evaluated by the author using six reproducible scenarios for correct and incorrect Stella programs. The evaluation confirmed the availability of the required data, synchronization with source ranges, bidirectional navigation, decomposition of typing steps, and highlighting of type conflicts. These results show that the prototype satisfies the functional criteria used in the evaluation. They do not prove that the extension improves learning outcomes, code readability, or debugging speed for other users because no controlled user study was conducted. A user study with objective task-completion measurements is therefore left for future work.
